---
  title: "Análises _preliminares_ dos marcadores COI e 12 no _metabarcoding_ de ictioplancton"
author: 
  - "Fonseca, DT; Hilário, OH; Carvalho, DC"
date: "07/12/2021"
output:
  html_document:
  code_download: yes
theme: flatly
toc: true
toc_depth: 4
toc_float: true
pdf_document: default
editor_options:
  chunk_output_type: console
---
  
  ```{r setup, include=FALSE}

```

***
  
  \pagebreak

# Bioinformática

## Comparação  

A partir dos resultados gerados em corridas independentes, iremos fazer comparações dos marcadores COI, MiFish & NeoFish.

### Load _R libs_ and set paths

```{r, eval=FALSE,echo=TRUE}
# 0 - load libraries ----
{
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(stringr)
  library(ggplot2)
  library(ggbreak)
  library(phyloseq)
  library(Biostrings)
  library(Matrix)
  library(ShortRead)
  library(dada2)
  library(DECIPHER)
  library(future)
  library(vegan)
}

#important paths
prjct_path <- "/home/heron/prjcts/fish_eDNA/ichthyo"

notes_path <- paste0(prjct_path,"/notes")

results_path <- paste0(prjct_path,"/results")

figs_path <- paste0(results_path,"/figs")

data_path <- paste0(prjct_path,"/data")


prjct_radical <- "ichthyoplancton_metabarcoding"
```

<br>
  
  ### Import and prepare  data
  
  ```{r, eval=FALSE,echo=TRUE}

all_IDs <- read.csv(file = paste0(data_path,"/","COI_e_12S_combinados.csv"),
                    header = TRUE,
                    check.names = FALSE,
                    na.strings=c("NA","NaN", "")) %>% 
  as_tibble()


# factorize variables for better ploting
all_IDs <- all_IDs %>% 
  mutate(Sample = as.factor(Sample),
         Primer = factor(Primer, levels = c("NeoFish","MiFish","COI")),
         Run = as.factor(Run))

```

<br>
  
  ### Exploratory plots
  
  #### Bar Plot
  
  #### Presence point plot
  

```{r, eval=FALSE,echo=TRUE}
#set colors for each primer

colors_norm <- c("#017504", #neo 
                 "#000791", #mif
                 "#cf9b02") #coi
scales::show_col(colors_norm)




plot_Id_species_per_sample <- 
  all_IDs %>% 
  ggplot(aes(y = `final ID`,
             x = Primer,
             col = Primer,
             size = `Relative abundance on sample (%)`,
             alpha = 0.5)) +
  geom_jitter(height = 0.01,width = 0.1) +
  scale_color_manual(values = colors_norm,
  ) +
  facet_grid(Sample ~ .,scales ='free_y', space ='free_y')


#save plot
ggsave(file = paste0(figs_path,"/",prjct_radical,"_identified_species.pdf"),
       plot = plot_Id_species_per_sample,
       device = "pdf",
       width = 30,
       height = 200,
       units = "cm",
       dpi = 600,
       limitsize = FALSE)

ggsave(file = paste0(figs_path,"/",prjct_radical,"_identified_species.png"),
       plot = plot_Id_species_per_sample,
       device = "png",
       width = 30,
       height = 200,
       units = "cm",
       dpi = 600,
       limitsize = FALSE)

```

<br>
  
  
  ### NMDS
  
  ```{r, eval=FALSE,echo=TRUE}
library(vegan)
# data(dune)
# decorana(dune)


all_IDs$Sample %>% unique()
all_IDs %>% colnames()

# class(dune)
#1- prepare data for entry in vegan ----

all_IDs_NMDS_tbl <- all_IDs %>% 
  filter(!(Sample %in% c("Cneg_PCR", "Neg1", "Neg2"))) %>%   #remove control samples
  select(c(Sample,Primer,Run,`final ID`,`Relative abundance on sample (%)`)) %>% #selecionar quais colunas aparecem
  unite("Sample_name", Sample, Primer,remove = FALSE ) %>% #cria a coluna sample_name com a uniao de sample e primer
  group_by(Sample_name,Sample,`final ID`,Primer,Run) %>%
  summarise(`Relative abundance on sample (%)` = sum(`Relative abundance on sample (%)`)) %>%
  pivot_wider(c(Sample_name,Sample,Primer,Run),names_from = `final ID` ,values_from = `Relative abundance on sample (%)`) %>%
  mutate_if(is.numeric, ~replace(., is.na(.), 0)) %>%
  mutate("Sample number" = 0) %>%
  ungroup()  %>%
  select(`Sample number`, 1:(ncol(.)-1))

#2- associate sample numbers to sample names ----
for (sample in 1:nrow(all_IDs_NMDS_tbl)) {
all_IDs_NMDS_tbl$`Sample number`[sample] <- sample
  # all_IDs_NMDS_tbl$`Sample number`[sample] <- all_IDs_NMDS_tbl$Sample[sample] %>% stringr::str_remove(pattern = "PP0|PP")
  }

#tirando as amostras da ecomol pra facilitar

# all_IDs_NMDS_tbl <- all_IDs_NMDS_tbl[all_ps_blst_vegan$Run %in% c("LGC_MiniSeq_1", "LGC_MiniSeq_2"),] 




colnames(all_IDs_NMDS_tbl)
hist(colSums(all_IDs_NMDS_tbl[,-c(1:5)]))
hist(rowSums(all_IDs_NMDS_tbl[,-c(1:5)]))
all_IDs_NMDS_tbl[,-c(1:5)]

all_IDs_NMDS_tbl %>% select(Sample_name, `Sample number`) %>% unique()
# all_ps_blst_vegan %>% select(`Sample number`, 1:(ncol(.)-1))

#3- create data.frame of species counts: rownames are Sample numbers ----

all_IDs_NMDS_df <- all_IDs_NMDS_tbl %>% 
  select(-c("Sample_name", "Sample", "Primer","Run")) %>% 
  select(base::sort(colnames(.))) %>% 
  as.data.frame() 

#4- name rows as Sample numbers and remove column ----
row.names(all_IDs_NMDS_df) <- all_IDs_NMDS_df$`Sample number`

all_IDs_NMDS_df <- all_IDs_NMDS_df %>% 
  select(-c(`Sample number`))


#5- 

all_ps_ord <- decorana(veg = all_IDs_NMDS_df)

all_ps_ord %>% summary()

all_ps_ord %>% str()

all_ps_ord$cproj


plot(all_ps_ord)
plot(all_ps_ord,type = "p")
plot(all_ps_ord,type = "c") 

points(all_ps_ord, display = "sites", cex = 0.8, pch=21, col="red", bg="yellow")
text(all_ps_ord, display = "sites", cex=0.7, col="blue")
text(all_ps_ord, display = "spec", cex=0.7, col="blue")



#6- NMDS analisys ----



# library(vegan)
# data(varespec)

#6a- Calculate distances ----
all_IDs_vg_dist <- vegdist(all_IDs_NMDS_df, method="bray")

vegan::scores(all_IDs_vg_dist)

# all_ps_vg_dist_metaMDS <- metaMDS(comm = all_ps_vg_dist, autotransform = FALSE) 
# actually autotransform = FALSE doesn't seem to change the results

# plot(all_ps_vg_dist_metaMDS)

# all_ps_vg_dist_metaMDS_2 <- metaMDS(comm = all_ps_vg_dist, distance = "bray", k =2)

# plot(all_ps_vg_dist_metaMDS_2)

#selecionar apenas espécies esperadas?

all_IDs_NMDS_df %>% ncol()
all_IDs_NMDS_df <- all_IDs_NMDS_df[,(colnames(all_IDs_NMDS_df) %in% expected_sps)]


all_IDs_NMDS_df %>% ncol()
all_IDs_vg_dist <- vegdist(all_IDs_NMDS_df, method="bray")

all_ps_ord <- decorana(veg = all_IDs_NMDS_df)

all_ps_ord %>% summary()

all_ps_ord %>% str()

all_ps_ord$cproj


plot(all_ps_ord)
plot(all_ps_ord,type = "p")
plot(all_ps_ord,type = "c") 
vegan::scores(all_IDs_vg_dist)




# all_IDs_NMDS_df[,(colnames(all_IDs_NMDS_df) %in% expected_sps)] %>% colnames()
# all_IDs_NMDS_df%>% colnames()



all_ps_vegan_ord_meta <- metaMDS(veg = all_IDs_NMDS_df, comm = all_IDs_vg_dist)
# actually autotransform = FALSE doesn't seem to change the results
plot(all_ps_vegan_ord_meta, type = "t")


all_ps_vegan_ord_meta %>% str()
all_ps_vegan_ord_meta

all_ps_vegan_ord_meta$stress

#6b- extract NMDS scores from results

all_vegan_meta <- (vegan::scores(all_ps_vegan_ord_meta) %>% 
                     tidyr::as_tibble(rownames = "Sample number")) %>% 
  mutate(`Sample number` = as.numeric(`Sample number`))
# all_vegan_meta <- as.data.frame(vegan::scores(all_ps_vegan_ord_meta))

#Using the scores function from vegan to extract the site scores and convert to a data.frame

# all_vegan_meta$`Sample number` <- rownames(all_vegan_meta) %>% as.numeric()  

# all_vegan_meta %>% left_join()# create a column of site names, from the rownames of data.scores

# all_vegan_meta <- all_vegan_meta  %>% as_tibble() # create a column of site names, from the rownames of data.scores

#7- bring NMDS scores to complete table

all_vegan_meta_tbl <- left_join(x = unique(all_IDs_NMDS_tbl[,c(1:5)]),
                                y = all_vegan_meta, by = "Sample number") %>% 
  mutate(Primer=factor(Primer,levels = c("NeoFish", "MiFish", "COI")),
         Sample = as.factor(Sample)) %>% 
  select(-c("Sample number"))





library(factoextra)
library(ggforce)




nmds_PLOT <- all_vegan_meta_tbl %>% 
  # filter(Run %in% c("LGC_MiniSeq_1", "LGC_MiniSeq_2")) %>% 
  ggplot(aes(x = NMDS1,y = NMDS2, col = Sample,
             shape = Primer,
             label = Sample,
             Group = Primer))+
  # stat_ellipse()+ 
  geom_point(size = 11)+
  theme_linedraw(base_size = 18) +
  theme(legend.position="bottom") +
  coord_fixed(ratio = 1) +
  # ggrepel::geom_label_repel(label.size = 0.8,size = 3,min.segment.length = 2) +
  # ggrepel::geom_text_repel(col="black",size = 3,min.segment.length = 2) +
  # scale_shape_manual() %>% 
  scale_color_manual(values = viridis::viridis(option = "turbo",n = 30, alpha = 1, begin = 0, end = 1, direction = 1)
                     # labels = c("NeoFish", "MiFish", "COI"), values = alpha(colour = colors_norm )
  ) +
  annotate(geom = "text",
           x=c(0.30),y=c(-0.3),label=c(paste0("Stress: ",all_ps_vegan_ord_meta$stress)),size=5)  +
  
  
  # ADD ggforce's ellipses
  ggforce::geom_mark_ellipse(inherit.aes = FALSE,
                             aes(x = NMDS1,y = NMDS2,
                                 group=`Sample`,
                                 label=`Sample`),
                             n = 100,
                             expand = 0.07,
                             label.fontsize = 20,con.cap = 0.1) 

nmds_PLOT
# 
# ggsave(file = "~/prjcts/fish_eDNA/sfjq/results/figs/SFJQ_NMDS.pdf",
#      plot = nmds_PLOT,
#      device = "pdf",
#      width = 14,
#      height =10,
#      dpi = 600)
# 
# ggsave(file = "~/prjcts/fish_eDNA/sfjq/results/figs/SFJQ_NMDS.svg",
#      plot = nmds_PLOT,
#      device = "svg",
#      width = 14,
#      height =10,
#      dpi = 600)


ggsave(file = paste0(figs_path,"/",prjct_radical,"_NMDS.pdf"),
       plot = nmds_PLOT,
       device = "png",
       width = 31,
       height =20,
       units = "cm",
       dpi = 300)

```

<br>
  
  **This is a partial report, intended to show the current state of analyses. Many procedures and conclusions might change as the pipeline evolves. If you notice errors/mistakes/typos, or have any suggestions, we would be glad to know. _heronoh@gmail.com_**
  